\chapter{Algorithmic and Quantitative Analysis}

This chapter explains the main algorithms used in the project and what they do
in practice. The formulas are here to show how the code works, not just to
list math on its own.

\section{Bit Encoding and Message Cost}
In early versions, each character is broken into 8 bits and transmitted as
separate UDP datagrams. If a message has length $L$ characters, then the number
of payload datagrams in pure bit mode is
\begin{equation}
N_{\text{bit}} = 8L + 8,
\end{equation}
where the additional $8$ corresponds to transmitting the null terminator as one
more byte.

That is why bit mode is slow. The number of datagrams grows very quickly.

\section{Parity-Based Error Detection}
For versions with parity, sender computes parity of one byte:
\begin{equation}
p = b_0 \oplus b_1 \oplus \cdots \oplus b_7,
\end{equation}
where $b_i$ is the $i$th bit of the character and $\oplus$ denotes XOR.

The receiver checks parity again and compares the result. If there is a
mismatch, that packet is marked as corrupted. Parity catches all odd bit errors
but not every even one, which is why later versions move to checksum.

\section{Stop-and-Wait Service Time}
In stop-and-wait, sender transmits one unit and waits for ACK before next send.
Approximate per packet cycle time is
\begin{equation}
T_{\text{cycle}} = T_{\text{tx}} + T_{\text{prop}} + T_{\text{proc}} + T_{\text{ack}}.
\end{equation}

Ideal throughput under stop-and-wait is therefore bounded by
\begin{equation}
\text{Throughput}_{\text{SAW}} \approx \frac{\text{payload per packet}}{T_{\text{cycle}}}.
\end{equation}

Even when channel capacity is high, waiting for each ACK keeps throughput low
if the round trip delay is not small.

\section{Checksum in TCP-like Versions}
In advanced versions, a one's complement style checksum is computed over key
header fields and payload bytes. In simple terms:
\begin{align}
S &= \sum \text{(header fields + payload bytes)}, \\
S' &= (S \bmod 2^{16}) + \left\lfloor\frac{S}{2^{16}}\right\rfloor, \\
\text{checksum} &= \sim S'.
\end{align}

At the receiver, checksum field is reset, recomputed, and compared. This gives
stronger detection than parity and supports multi byte payload packets.

\section{Sliding Window Efficiency}
With sender window size $W$, up to $W$ packets can be in flight. In an ideal
pipeline, normalized channel utilization can be approximated as
\begin{equation}
U \approx \min\left(1, \frac{W}{1 + 2a}\right),
\end{equation}
where $a = T_{\text{prop}}/T_{\text{tx}}$.

This is why a larger window increases throughput until the path is fully used.

\section{Chunking Gain}
If payload per packet increases from 1 byte to $C$ bytes (for example,
$C=8$), the number of data packets for message length $L$ becomes
\begin{equation}
N_{\text{chunk}} = \left\lceil\frac{L}{C}\right\rceil.
\end{equation}

Compared with one-byte payloads, the packet count drops by about
\begin{equation}
\text{Gain factor} \approx C,
\end{equation}
ignoring control packets and retransmissions.

\clearpage
\section{Adaptive RTO Estimation}
Versions 12 and above use smoothed RTT estimation and variation based timeout.
Given sample RTT $R$, smoothed RTT ($SRTT$) and RTT variation ($RTTVAR$) are
updated as follows:
\begin{align}
RTTVAR &\leftarrow (1-\beta)\,RTTVAR + \beta\,|SRTT - R|, \\
SRTT   &\leftarrow (1-\alpha)\,SRTT + \alpha\,R.
\end{align}

Then retransmission timeout is estimated by
\begin{equation}
RTO = SRTT + K \cdot RTTVAR,
\end{equation}
and bounded in implementation between predefined minimum and maximum values.

This model is aligned with practical transport engineering guidance and avoids
both aggressive and overly delayed retransmissions.

\section{Congestion Window Dynamics}
In congestion controlled versions, sender maintains congestion window $cwnd$ and
slow-start threshold $ssthresh$.

\subsection{Slow Start}
When $cwnd < ssthresh$, growth is fast (approximately one MSS equivalent unit
per ACK, resulting in near exponential growth per RTT in simplified models).

\subsection{Congestion Avoidance}
When $cwnd \ge ssthresh$, growth becomes gradual, approximating additive
increase.

\subsection{Loss Response}
On timeout event:
\begin{align}
ssthresh &\leftarrow \max\left(2, \frac{cwnd}{2}\right), \\
cwnd &\leftarrow 1.
\end{align}

On triple duplicate ACK in fast retransmit logic:
\begin{align}
ssthresh &\leftarrow \max\left(2, \frac{cwnd}{2}\right), \\
cwnd &\leftarrow ssthresh + 3.
\end{align}

After successful recovery, window is deflated near threshold level. This avoids
fully restarting from 1 in many recoverable loss cases.

\section{Latency Components}
Observed transfer completion time includes multiple components:
\begin{equation}
T_{\text{total}} = T_{\text{setup}} + T_{\text{data}} + T_{\text{retrans}} + T_{\text{teardown}}.
\end{equation}

Where:
\begin{enumerate}
    \item $T_{\text{setup}}$ includes handshake steps,
    \item $T_{\text{data}}$ includes normal in order transfer,
    \item $T_{\text{retrans}}$ includes recovery delay from loss/corruption,
    \item $T_{\text{teardown}}$ includes connection close signaling.
\end{enumerate}

This decomposition helps explain why different versions show different
completion times for the same message length.

\section{Reliability Metrics}
Useful metrics for experiment reporting are defined below.

\begin{align}
\text{Delivery Success Rate} &= \frac{\text{Successfully delivered messages}}{\text{Messages attempted}}, \\
\text{Retransmission Ratio} &= \frac{\text{Retransmitted packets}}{\text{Original packets}}, \\
\text{Goodput} &= \frac{\text{Useful payload bits delivered}}{T_{\text{total}}}.
\end{align}

These metrics give a balanced view across correctness and efficiency.

\section{A Practical Theorem-like Observation}
\begin{theorem}
Increasing the size of sender's window increases throughput up to the
saturation point and, thereafter, throughput is limited.
\end{theorem}

\begin{Proof}
Increasing the number of packets in transmission reduces total time sender
waits on acknowledgement before sending the next packet in sequence. The
larger the packet window, the more packets can be transmitted before sender has
to wait on an acknowledgement; and, therefore, the fewer times the sender will
be idle. However, once the number of in transit packets is equal to the
product of the time it takes for a packet to travel from sender to receiver
and the receiver's ability to process packets (e.g. memory); no increase in
sender's window will produce any additional packets delivered. Thus, when
there are lost packets in error, a larger sending window will produce a larger
error recovery burden unless recipient uses selective acknowledgment.
Consequently, throughput will plateau once sender reaches maximum size of
window.
\end{Proof}

\clearpage
\section{Interpretation for This Project}
The formulas in this chapter explain observed behavior from level progression:

\begin{enumerate}
    \item versions v0--v2 have very high packet overhead,
    \item v5--v7 improve correctness but remain latency bound,
    \item v9 and above improve utilization through pipelining,
    \item v10 and above reduce packet count through chunking,
    \item v12 and above improve timing decisions through adaptive RTO,
    \item v13 and above balance aggressiveness and stability using congestion
    control logic and recovery aware retransmission behavior,
    \item v15--v24 extend the same transport ideas with richer TCP-like
    refinements, including timestamp and recovery oriented control.
\end{enumerate}

This quantitative interpretation supports the implementation narrative in the
previous chapter and sets context for the comparative results in the next
chapter.