\chapter{Extended Technical Appendix}

This appendix provides additional implementation depth for the TCP/IP-inspired
UDP project. It is written to support evaluation, viva discussion, and future
continuation of the codebase.

\section{Appendix Purpose}
The main chapters describe architecture and results at report level. This
appendix goes deeper into operational details:

\begin{enumerate}
    \item detailed behavior of sender and receiver loops,
    \item practical debug checklist used during development,
    \item experiment script style and repeatable run plan,
    \item level-specific observations and pitfalls,
    \item extension blueprint for future cohorts.
\end{enumerate}

\section{Complete Build and Run Workflow}
The project uses plain GCC and does not depend on external networking
frameworks. A standard run for one level follows the same sequence.

\subsection{Compile}
For version $N$:
\begin{enumerate}
    \item \texttt{gcc -std=c99 -o receiver\_N receiver\_N.c}
    \item \texttt{gcc -std=c99 -o sender\_N sender\_N.c}
\end{enumerate}

\subsection{Execute}
\begin{enumerate}
    \item Open terminal 1 and run \texttt{./receiver\_N}
    \item Open terminal 2 and run \texttt{./sender\_N}
    \item Enter test message in sender terminal.
\end{enumerate}

\subsection{Operational Notes}
\begin{enumerate}
    \item Always start receiver first so bind/listen path is active.
    \item Keep one active sender-receiver pair on port 9090 at a time.
    \item If process hangs after experiments, terminate old binaries and rerun.
\end{enumerate}

\section{Debug Checklist Used in This Project}
The following checklist was used repeatedly and is useful for future maintenance.

\begin{enumerate}
    \item Verify sender destination address and port.
    \item Verify receiver bind success and error handling.
    \item Confirm packet field initialization before checksum computation.
    \item Confirm checksum recomputation resets checksum field to zero.
    \item Confirm sender base and next\_to\_send pointers satisfy
    $base \le next\_to\_send$.
    \item Confirm ACK handling updates sender window exactly once.
    \item Confirm duplicate packet handling does not duplicate delivery.
    \item Confirm timeout value is in safe range and retry cap prevents
    infinite loops.
    \item Confirm FIN handshake executes and sockets are closed gracefully.
\end{enumerate}

\clearpage
\section{Detailed Behavior: Sender Core Loop}
Across advanced versions, sender logic follows a common event-driven pattern:

\begin{enumerate}
    \item Initialize socket, destination, timeout controls.
    \item Perform handshake if protocol level requires session setup.
    \item Segment message into packets/chunks.
    \item Transmit packets while keeping in-flight count below window limit.
    \item Wait for ACKs and slide window accordingly.
    \item Detect timeout or duplicate ACK events and trigger retransmission
    policy.
    \item Close connection using FIN logic.
\end{enumerate}

\subsection{State Variables}
The most influential sender variables in high-level versions are:

\begin{enumerate}
    \item \texttt{base}: oldest unacknowledged sequence.
    \item \texttt{next\_to\_send}: next sequence eligible for send.
    \item \texttt{window\_size}: configured or dynamic send limit.
    \item \texttt{rto\_ms}: retransmission timeout in milliseconds.
    \item \texttt{cwnd} and \texttt{ssthresh}: congestion-control state.
    \item duplicate ACK counters per acknowledgement number.
\end{enumerate}

\clearpage
\section{Detailed Behavior: Receiver Core Loop}
Advanced receiver implementations use predictable stages:

\begin{enumerate}
    \item Validate packet checksum and flags.
    \item Check sequence relation with \texttt{expected\_seq}.
    \item For in-order data, deliver immediately and advance expectation.
    \item For out-of-order but in-window data, buffer and ACK appropriately.
    \item For duplicate data, re-ACK to help sender convergence.
    \item Process FIN and release resources cleanly.
\end{enumerate}

\subsection{Receiver Buffering in Selective-Repeat Style}
Selective-repeat-like receiver behavior uses parallel arrays for packet
presence, payload storage, and payload length. This avoids data loss during
reordering and allows in-order flush once missing sequence arrives.

\section{Version-by-Version Pitfall Notes}
This section captures practical issues found during progressive development.

\subsection{v0--v2 Pitfalls}
\begin{enumerate}
    \item Mistake in bit ordering can invert character reconstruction.
    \item Missing null terminator logic causes receiver to block indefinitely.
    \item Delay control via \texttt{usleep} is environment-dependent.
\end{enumerate}

\subsection{v3--v7 Pitfalls}
\begin{enumerate}
    \item Parity mismatch handling must be deterministic on both ends.
    \item ACK/NAK message format ambiguity can break sender wait logic.
    \item Retry loops must cap attempts to avoid infinite retransmit cycles.
\end{enumerate}

\subsection{v8--v10 Pitfalls}
\begin{enumerate}
    \item Handshake flags must not overlap incorrectly.
    \item Checksum must be recomputed after all payload bytes are assigned.
    \item Sliding window pointer arithmetic is prone to off-by-one errors.
\end{enumerate}

\subsection{v11--v14 Pitfalls}
\begin{enumerate}
    \item Out-of-order buffer flush must preserve sequence continuity.
    \item Adaptive RTO can oscillate without stable RTT sampling.
    \item Duplicate ACK counters must reset correctly after state transitions.
    \item Fast recovery exits can be mishandled if ACK interpretation is vague.
\end{enumerate}

\section{Experiment Design Template}
A repeatable experiment protocol is essential for fair comparison. The template
used in this project is shown below.

\begin{table}[h]
\centering
\captionabove{Standard experiment template used for each protocol version.}
\renewcommand{\arraystretch}{1.18}
\begin{tabular}{|p{3.5cm}|p{9.4cm}|}
\hline
\textbf{Field} & \textbf{Description} \\
\hline
Version & sender\_N / receiver\_N pair \\
\hline
Message length & Short (20 chars), medium (60 chars), long (95 chars) \\
\hline
Fault setting & No fault / delay / simulated loss / duplicate packet scenario \\
\hline
Runs per case & Minimum 10 runs for stable average \\
\hline
Metrics & Completion time, retransmissions, ACK count, success/failure \\
\hline
Notes & Any abnormal behavior or timeout event logs \\
\hline
\end{tabular}
\label{tab:experiment-template}
\end{table}

\clearpage
\section{Sample Observation Narrative}
The following narrative style was used to interpret test runs.

\begin{displayquote}
In level 9 under moderate induced delay, sender experienced repeated timeout at
window base. Retransmission from base restored progress, but overall completion
time increased significantly compared to delay-free run. Receiver ACK pattern
confirmed cumulative behavior.
\end{displayquote}

\begin{displayquote}
In level 12 under variable delay, adaptive RTO reduced false timeout events
after initial RTT stabilization. Retransmission ratio decreased compared to
fixed-timeout configuration.
\end{displayquote}

\begin{displayquote}
In level 14 with synthetic packet gap, triple duplicate ACK condition triggered
fast retransmit before timeout expiry. Recovery was visibly faster than
timeout-driven recovery path.
\end{displayquote}

\section{Illustrative State Machine View}
\vspace{2\baselineskip}
\begin{figure}[H]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
    stA/.style={draw, rounded corners=4pt, thick, minimum width=4.3cm, minimum height=1.15cm, align=center, fill=cyan!14, draw=cyan!55!black},
    stB/.style={draw, rounded corners=4pt, thick, minimum width=4.3cm, minimum height=1.15cm, align=center, fill=yellow!18, draw=orange!65!black},
    stC/.style={draw, rounded corners=4pt, thick, minimum width=4.3cm, minimum height=1.15cm, align=center, fill=green!14, draw=green!45!black},
    arr/.style={-{Latex[length=3.2mm]}, very thick, draw=black!82},
    lab/.style={fill=white, inner sep=1.2pt, font=\footnotesize}
]
\node[stA] (idle) at (0,0) {Idle};
\node[stB] (syn) at (5.8,0) {Handshake};
\node[stC] (send) at (11.6,0) {Send Window};
\node[stC] (wait) at (17.4,0) {Wait ACK};
\node[stA] (fin) at (23.2,0) {Teardown};
\node[stB] (retx) at (14.5,-2.9) {Fast Retransmit};

\draw[arr] (idle.east) -- (syn.west);
\draw[arr] (syn.east) -- (send.west);
\draw[arr] (send.east) -- (wait.west);
\draw[arr] (wait.east) -- node[above,lab]{all data acked} (fin.west);
\draw[arr] (wait.south) -- node[right,lab]{timeout / dup ACKs} (retx.north);
\draw[arr] (retx.west) to[bend right=18] node[below,lab]{retry path} (send.south);
\end{tikzpicture}%
}%

\caption{Simplified sender-side operational state flow in advanced versions.
}
\label{fig:appendix-state-machine}
\end{figure}

\section{Protocol Trace Formatting Recommendation}
\vspace{0.8\baselineskip}

\begin{figure}[H]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
    fld/.style={draw, rounded corners=4pt, thick, minimum width=4.6cm, minimum height=1.25cm, align=center, fill=teal!10, draw=teal!60!black, font=\small},
    eventfld/.style={draw, rounded corners=4pt, thick, minimum width=6.1cm, minimum height=1.25cm, align=center, fill=teal!10, draw=teal!60!black, font=\small},
    arr/.style={-{Latex[length=3.1mm]}, very thick, draw=black!82}
]
\node[fld] (f1) at (0,0) {[timestamp]};
\node[fld] (f2) at (5.5,0) {[role=sender|receiver]};
\node[eventfld] (f3) at (12.2,0) {[event=send|recv|ack|timeout|retransmit]};

\node[fld] (f4) at (0,-2.6) {[seq=x] [ack=y]};
\node[fld] (f5) at (5.5,-2.6) {[cwnd=z] [rto=ms]};
\node[fld] (f6) at (12.2,-2.6) {[note=text]};

\draw[arr] (f1.east) -- (f2.west);
\draw[arr] (f2.east) -- (f3.west);
\draw[arr] (f1.south) -- (f4.north);
\draw[arr] (f2.south) -- (f5.north);
\draw[arr] (f3.south) -- (f6.north);
\draw[arr] (f4.east) -- (f5.west);
\draw[arr] (f5.east) -- (f6.west);
\end{tikzpicture}%
}%

\caption{Recommended structured protocol-trace fields for automated analysis.
}
\label{fig:trace-format-diagram}
\end{figure}

\section{Suggested Quantitative Extensions}
Future project groups can convert this implementation into a mini benchmark
suite by adding automated scripts. Useful extensions include:

\begin{enumerate}
    \item throughput vs message size curve per version,
    \item retransmission ratio vs induced packet-loss probability,
    \item completion time distribution under delay jitter,
    \item fairness comparison under two parallel sender flows,
    \item sensitivity of adaptive RTO to \(\alpha\) and \(\beta\) choices.
\end{enumerate}

\section{Security and Reliability Perspective}
The project intentionally focuses on reliability and congestion behavior, not
security. However, in practical deployments, integrity and authenticity both
matter. Checksum detects random corruption but does not prevent malicious packet
modification. A production extension would include:

\begin{enumerate}
    \item message authentication,
    \item optional encryption,
    \item replay protection,
    \item session identity validation.
\end{enumerate}

These are orthogonal to transport logic but critical for real systems.

\clearpage
\section{Code Organization Recommendations}
As code complexity grows, modularization becomes important. Recommended
refactoring pattern:

\begin{enumerate}
    \item \texttt{packet.h/.c} for packet structs and checksum utilities,
    \item \texttt{timer.h/.c} for RTT/RTO and timestamp helpers,
    \item \texttt{window.h/.c} for send/receive window operations,
    \item \texttt{congestion.h/.c} for cwnd and recovery transitions,
    \item \texttt{main\_sender.c} and \texttt{main\_receiver.c} for orchestration.
\end{enumerate}

This separation reduces duplication across version files.

\section{Reproducibility Notes}
To ensure reproducibility of results in academic evaluation:

\begin{enumerate}
    \item record compiler version and build flags,
    \item record operating system and kernel details,
    \item isolate background network-intensive applications,
    \item run each experiment multiple times,
    \item report average and spread (minimum/maximum or standard deviation).
\end{enumerate}

Without reproducibility controls, protocol comparisons may appear inconsistent.

\section{Expanded Reflection on Learning Outcomes}
This project created strong practical learning in four dimensions.

\subsection{Networking Concepts}
Ideas such as reliable delivery, sequence management, and ACK policies became
much clearer after implementing them manually.

\subsection{Systems Programming}
Low-level C socket work, timeout configuration, and careful memory handling
improved confidence in systems programming discipline.

\subsection{Algorithmic Thinking}
Adaptive timeout and congestion behavior required state-machine thinking and
careful event response logic.

\subsection{Technical Communication}
The staged codebase enabled structured report writing, where each chapter and
section maps naturally to a protocol capability.

\section{Expanded Testing Scenarios and Interpretation}
Recommended scenarios include clean-channel baseline, delayed ACK,
packet-loss injection, out-of-order arrival, duplicate ACK bursts, and
near-limit message size tests. Together, these cases expose correctness,
recovery speed, timeout behavior, and congestion adaptation quality.

\section{Evaluation Rubric for Academic Assessment}
The following rubric can be used during project review.

\begin{table}[h]
\centering
\captionabove{Suggested assessment rubric for implementation quality.}
\renewcommand{\arraystretch}{1.18}
\begin{tabular}{|p{3.3cm}|>{\centering\arraybackslash}p{2.1cm}|p{7.3cm}|}
\hline
\textbf{Criterion} & \textbf{Weight} & \textbf{Evaluation Focus} \\
\hline
Correctness & 30\% & Data integrity checks, ordered delivery, clean teardown \\
\hline
Reliability & 20\% & Behavior under loss, timeout and retransmission policy quality \\
\hline
Efficiency & 15\% & Throughput trend improvement from low to high levels \\
\hline
Code Quality & 15\% & Clarity, modularity, safety checks, variable naming \\
\hline
Analysis & 10\% & Metric reporting and explanation quality \\
\hline
Documentation & 10\% & Diagram quality, chapter structure, reproducibility notes \\
\hline
\end{tabular}
\label{tab:assessment-rubric}
\end{table}

\clearpage
\section{Final Reflection}
From a learning viewpoint, the strongest contribution of this project is the
clear mapping between theory and implementation. Concepts often memorized in
networking courses become operational behaviors that can be observed directly in
terminal traces. The staged versions form a practical curriculum in protocol
engineering: encoding, integrity checks, reliability loops, pipelining,
adaptation, and congestion response.

This project also demonstrates that protocol design is fundamentally about
trade-offs. Stronger reliability adds state. Higher throughput adds complexity.
Faster recovery requires careful event interpretation. Better adaptability needs
stable measurement. Understanding and balancing these trade-offs is the core
skill developed through this work.

\section{Appendix Summary}
This appendix extends the main report with operational detail, implementation
insights, and actionable next steps. Together with the main chapters, it
provides a complete technical narrative from basic UDP communication to
TCP-inspired reliability and congestion control behavior in C.TCP-inspired transport control through deep learning exercises.
